---
title: "ST 590 Project 1"
author: "Amanda, adil & Brian"
date: "September 28, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries
Test pushing Rmarkdown file. Content to be filled in. 
```{r, warning = FALSE, message = FALSE}
library(tidyverse)
#requiring more packages that will be helpful later to handle XML data
library(XML)
library(plyr)
library(dplyr)
```


## Reading in XML data

I found a website with many public xml datasets, some of them contain  very ineteresting information. (http://aiweb.cs.washington.edu/research/projects/xmltk/xmldata/www/repository.html)
Mondial is an XML dataset from World geographic database integrated from the CIA World Factbook, the International Atlas, and the TERRA database among other sources.
The data is made public and able to download from washington.edu datasets repository online.
I was able to parsethe XML data - see code below but I'mstill trying to get the data into a DF.

```{r Mondial}
mondialParse <- xmlParse("http://aiweb.cs.washington.edu/research/projects/xmltk/xmldata/data/mondial/mondial-3.0.xml")
class(mondialParse) # Testing to make sure that R recognizes the XML data

```
```{r MondialDF}
#commenting this code for now until we figure out how to get XML into a DF 
#mondialDF <- xmlToDataFrame(mondialParse)

```

## Data wrangling

## Variable Creation

## Aggregation summary
