---
title: "ST 590 Project 1"
author: "Amanda, adil & Brian"
date: "September 28, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries
```{r, warning = FALSE, message = FALSE}
library(tidyverse)
#requiring more packages that will be helpful later to handle XML data
library(XML)
library(plyr)
library(dplyr)
```


## Reading in XML data

We found a website with many public xml datasets, some of them contain very ineteresting information.<br> (http://aiweb.cs.washington.edu/research/projects/xmltk/xmldata/www/repository.html)
Mondial is an XML dataset from World geographic database integrated from the CIA World Factbook, the International Atlas, and the TERRA database among other sources.<br>
The data is made public and able to download from washington.edu datasets repository online.

```{r Mondial}

mondialParse <- xmlParse("http://aiweb.cs.washington.edu/research/projects/xmltk/xmldata/data/mondial/mondial-3.0.xml")
class(mondialParse) # Testing to make sure that R recognizes the XML data

```

## Transforming XML into a dataframe

```{r xmlToDF}

monialDF <- ldply(xmlToList("http://aiweb.cs.washington.edu/research/projects/xmltk/xmldata/data/mondial/mondial-3.0.xml"), data.frame)
view(mondialDF) # Tried the code above but result in an error: arguments imply differing number of rows: 1, 4, 2, 14

```


## Variable Creation

## Aggregation summary
